---
title: "linreg-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{linreg-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lab4linreg)
```

The lab4linreg package performs linear regression using QR decomposition and has a number of useful functions for accessing and plotting the results.

```{r, message = FALSE, warning = FALSE, fig.align='center', fig.height = 4, fig.width = 6}
linreg_mod <- linreg$new(formula = Petal.Length ~ Species, data=iris)
linreg_mod$plot()
```

### print
print() prints out the coefficients and coefficient names, similar as done by the lm class.

```{r print}
linreg_mod
```
### Resid

restid() returns the vector of residuals.

```{r resid}
linreg_mod$resid()

```
### Pred

pred() returns the predicted values

```{r pred}

linreg_mod$pred()

```
### Coef
Tcoef() returns the coefficients as a named vector

```{r coef}
linreg_mod$coef()
```


### Summary

summary() returns a similar printout as printed for lm objects, but you only need to present
the coefficients with their standard error, t-value and p-value as well as the estimate of ˆσ and the degrees
of freedom in the model.

```{r Summary}

linreg_mod$summary()

```
